Left-linear rules in adhesive categories

### 2024-02-11 icalp/fusione.tex
abstract

\begin{abstract}
  When two sequences of steps executed by a computational device can
  be considered equivalent?  This is a relevant question if one
  considers concurrent and distributed systems, since the independence
  of two steps means that they could be executed on any order, and
  potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, which
  allows to properly extend the formalisms that can be modelled in the
  adhesive framework, and for which we prove a Church-Rosser property.
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  The canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear
  rewriting systems.  Additionally, a fundamental characterisation of
  the notion of independence fails when moving to left-linear
  rewriting systems. This requires a substantial revision of the
  theory as in the linear case, a number of basic properties,
  essential in the development of the theory of concurrency, were
  derived from such a characterisation.
\end{abstract}

@@

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, this equivalence 
captures when steps are causally unrelated and thus could be executed in any 
order, and possibly concurrently. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules which have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
As noteworthy examples, we have 
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

>>>>

\begin{abstract}
  When two sequences of steps executed by a computational device can
  be considered equivalent?  This is a relevant question if one
  considers concurrent and distributed systems, since the independence
  of two steps means that they could be executed in any order, and
  potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  In fact, the standard characterisation of independence does not hold anymore,
  and this requires a substantial revision of the 
  theory for the linear case, since a number of basic properties,
  essential in the development of a theory of concurrency, were
  derived from such a description.
  %
  The paper introduce a novel characterisation of independence, for which we prove a Church-Rosser property,
  properly extending the class of formalisms 
  that can be modelled in the adhesive framework. 
\end{abstract}

@@

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when steps are causally unrelated and thus could be executed in any 
order, and possibly concurrently. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules which have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
As noteworthy examples, we have 
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.
